Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo

### 2012-02-16 Paper/draft5_MCMCsmoother.tex
Final draft for Simon to proofread.

%% Packages
\usepackage{cite}
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  \graphicspath{{../pdf/}{../jpeg/}}
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  \usepackage[dvips]{graphicx}
  \graphicspath{{../eps/}}
  \DeclareGraphicsExtensions{.eps}
\fi
\usepackage[cmex10]{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
%\usepackage{stfloats}
\usepackage{url}

@@

The particle filter is summarised below.%in algorithm~\ref{alg:PF}.

%\begin{algorithm}
\begin{algorithmic}
	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
	\FOR{$k = 1 \dots K$}
 		\FOR{$i = 1 \dots N_F$}
 			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 			\STATE Discard $x_{k-1}^{(i)}$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
%\caption{Particle filter algorithm}
%\label{alg:PF}
%\end{algorithm}

@@

The filter-smoother is summarised below.%in algorithm~\ref{alg:FS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
 	\FOR{$k = 1 \dots K$}
 		\FOR{$i = 1 \dots N_F$}
 			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
 \end{algorithmic}
%\caption{Filter-smoother algorithm}
%\label{alg:FS}
%\end{algorithm}

@@

The procedure may be repeated to produce as many samples from the joint smoothing distribution as required. This backward-resampling smoother is summarised below.% in algorithm~\ref{alg:DBRS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
		\FOR{$k = K-1 \dots 1$}
			\FOR{$j = 1 \dots N_F$}
				\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
			\ENDFOR
			\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{Direct backward-resampling smoother algorithm}
%\label{alg:DBRS}
%\end{algorithm}

@@

The complete MCMC backward-resampling smoother is summarised below.% in algorithm~\ref{alg:MCMC-BRS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BRS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-resampling smoother algorithm}
%\label{alg:MCMC-BRS}
%\end{algorithm}

@@

@@

The proposed new method proceeds in a similar manner to the MCMC backward-resampling algorithm. However, this time $x_{1:k-1}$, is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows:

\begin{IEEEeqnarray}{rCl}
q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) & = & q(x_{1:k-1}|y_{1:k-1}) q(x_{k}|x_{k-1}, \tilde{x}_{k+1}, y_{k}) \nonumber \\
                                    & = & \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1}) q(x_{k}|x_{k-1}^{(j)}, \tilde{x}_{k+1}, y_{k})  .
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) & =       & p(x_{1:k}|\tilde{x}_{k+1}, y_{1:k}) \nonumber \\
                                    & \propto & p(\tilde{x}_{k+1}|x_k) p(x_k|x_{k-1}) p(y_k|x_x) p(x_{1:k-1}|y_{1:k-1})
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
\alpha &=& \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \nonumber \\
 &=& \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \nonumber \\
 \IEEEeqnarraymulticol{3}{l}{\frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } } .
\end{IEEEeqnarray}

The complete MCMC backward-sampling smoother is summarised below.% in algorithm~\ref{alg:MCMC-BSS}.

%\begin{algorithm}
 \begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state hisory, $x_{1:k-1}^{(i)*} \sim \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim q(x_k|x_{k-1}^{*},\tilde{x}_{k+1},y_k)$.
				\STATE With probability $\alpha = \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-sampling smoother algorithm}
%\label{alg:MCMC-BSS}
%\end{algorithm}

@@

The performance of the new smoothers was tested on a simple 2D tracking model. The trajectory of a single target was simulated using linear-Gaussian dynamics and observed via a nonlinear radar-type measurement model.

\begin{IEEEeqnarray}{rCl}
\underbrace{\begin{bmatrix}x_k \\ y_k \\ \dot{x}_k \\ \dot{y}_k\end{bmatrix}}_{\mathbf{x}_k} &=& \underbrace{\begin{bmatrix}I_2 & \Delta t I_2 \\ 0_2 & I_2\end{bmatrix}}_{\mathbf{A}} \underbrace{\begin{bmatrix}x_{k-1} \\ y_{k-1} \\ \dot{x}_{k-1} \\ \dot{y}_{k-1}\end{bmatrix}}_{\mathbf{x}_{k-1}} + \underbrace{\begin{bmatrix}w_{1,k} \\ w_{2,k} \\ w_{3,k} \\ w_{4,k}\end{bmatrix}}_{\mathbf{w}_k} \\
\underbrace{\begin{bmatrix}b_k \\ r_x\end{bmatrix}}_{\mathbf{y}_k} &=& \begin{bmatrix}\tan^{-1}(y_k / x_k) \\ \sqrt{x_k^2+y_k^2} \end{bmatrix} + \underbrace{\begin{bmatrix}v_{1,k} \\ v_{2,k}\end{bmatrix}}_{\mathbf{v}_k}
\end{IEEEeqnarray}

The time between successive time steps is $\Delta t$. $I_2$ and $0_2$ denote the $2 \times 2$ identity and zero matrices. The random variables $\mathbf{w}_k$ and $\mathbf{v}_k$ have a zero-mean Gaussian distribution with covariance matrices $\mathbf{Q}$ and $\mathbf{R}$ respectively.

@@

The algorithms were all used to generate 100 particles from the joint smoothing distribution, using particle filters with 100 particles. Where appropriate, proposal weights were chosen to be equal to filtering weights, i.e. $v_k^{(j)} = w_k^{(j)}$. The particle filter and MCMC-BSS both use ``optimal importance distributions'' (see e.g. \cite{Cappe2007}) with linearisations of the observation model where appropriate.

10 realisations of 500 time steps were simulated from the model and the algorithms tested on each one. Results are shown here for high and low observation noise cases. For case 1, the observation variances were set to $\sigma_B = \pi/720$ and $\sigma_R = 0.1$. For case 2, $\sigma_B = \pi/36$ and $\sigma_R = 100$. Other model settings are $\Delta t = 1$, $\sigma_P = 1$, and $x_0 = [-100, 50, 10, 0]^T$. Example tracjectories and observations from the two cases are shown in figure~\ref{fig:example_trajectories}.

@@

Tables~\ref{tab:case1_performance} and~\ref{tab:case1_performance} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms with the two parameter cases, along with running times (excluding the time for filtering) and the mean number of unique particles at each time instant. The trade-off between speed and error is shown in figures~\ref{fig:case1_rmse_vs_time} and~\ref{fig:case2_rmse_vs_time}.

@@

The MCMC-BRS smoother error approaches that of the DBRS as the number of MCMC steps, $M$, increases. However, for small $M$, the error is only very slightly larger than that of the DBRS for very low running times. For a particular application, it is possible to adjust $M$ to control the trade-off between error and speed.

@@

\section{Conclusion} \label{sec:conclusions}
Two new algorithms have been presented for drawing particles from the joint smoothing distribution using MCMC. The MCMC schemes have a degree of flexibility in the choice of chain length, which controls a trade-off between accuracy and running time.

@@

The second algorithm samples new values of the state from a proposal distribution. Although slower than the resampling version for a fixed length of Markov chain, the error performance is reduced relative to the other smoothers.

>>>>

%% Packages
\usepackage{cite}
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  \graphicspath{{../pdf/}{../jpeg/}}
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  \usepackage[dvips]{graphicx}
  \graphicspath{{../eps/}}
  \DeclareGraphicsExtensions{.eps}
\fi
\usepackage[cmex10]{amsmath}
%\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
%\usepackage{stfloats}
\usepackage{url}

@@

The particle filter is summarised in Fig.~\ref{alg:PF}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 			\STATE Discard $x_{k-1}^{(i)}$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Particle filter algorithm}
\label{alg:PF}
\end{figure}
%\end{algorithm}

@@

The filter-smoother is summarised in Fig.~\ref{alg:FS}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
 	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
 	\FOR{$k = 1$ \TO $K$}
 		\FOR{$i = 1$ \TO $N_F$}
 			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
}}
\caption{Filter-smoother algorithm}
\label{alg:FS}
\end{figure}
%\end{algorithm}

@@

The procedure may be repeated to produce as many samples from the joint smoothing distribution as required. This backward-resampling smoother is summarised in Fig.~\ref{alg:DBRS}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
 	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
		\FOR{$k = K-1$ \TO $1$}
			\FOR{$j = 1$ \TO $N_F$}
				\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
			\ENDFOR
			\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{Direct backward-resampling smoother algorithm}
\label{alg:DBRS}
\end{figure}
%\end{algorithm}

@@

The complete MCMC backward-resampling smoother is summarised in Fig.~\ref{alg:MCMC-BRS}.

%\begin{algorithm}
\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1$ \TO $1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1$ \TO $M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BRS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{MCMC backward-resampling smoother algorithm}
\label{alg:MCMC-BRS}
\end{figure}
%\end{algorithm}

@@

Smoothers based on backward-resampling, both the direct sampler of \cite{Godsill2004} and the MCMC version outlined in the previous section, are limited by the fact that states are only selected from those which appear in the filtering approximation. If there is a significant difference between the filtering and smoothing densities, then the the particles are unlikely to be in the right locations to represent the smoothing distribution well, and the resulting approximation will be poor.

@@

The proposed new method proceeds in a similar manner to the MCMC backward-resampling algorithm. However, this time only $x_{1:k-1}$ is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows:

\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ q(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = q(x_{1:k-1}|y_{1:k-1}) q(x_{k}|x_{k-1}, \tilde{x}_{k+1}, y_{k}) } \nonumber \\
                                    & = & \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1}) q(x_{k}|x_{k-1}^{(j)}, \tilde{x}_{k+1}, y_{k})  .
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(x_{1:k}|\tilde{x}_{k+1}, y_{1:K}) = p(x_{1:k}|\tilde{x}_{k+1}, y_{1:k}) } \nonumber \\
                                    & \propto & p(\tilde{x}_{k+1}|x_k) p(x_k|x_{k-1}) p(y_k|x_x) p(x_{1:k-1}|y_{1:k-1})
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
\alpha &=& \min \Bigg[ 1, \frac{ \hat{p}(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_{1:k}^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_{1:k}^{*}|\tilde{x}_{k+1}, y_{1:K}) } \Bigg] \nonumber \\
 &=& \min \Bigg[ 1, \frac{ w_{k-1}^{*} p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) }{ w_{k-1}^{(m)} p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) } \nonumber \\
 & & \times \frac{ \tilde{v}_{k-1}^{(m)} q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ \tilde{v}_{k-1}^{*} q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } \Bigg] \label{eq:MCMC-BSS_ap}
% \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \nonumber \\
% \IEEEeqnarraymulticol{3}{l}{\frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) } } .
\end{IEEEeqnarray}

As noted in \cite{Fearnhead2010}, the optimal choice of proposal is the conditional posterior state distribution,

\begin{IEEEeqnarray}{rCl}
q(x_k|x_{k-1},x_{k+1},y_k) & = & p(x_k|x_{k-1},x_{k+1},y_k)  .
\end{IEEEeqnarray}

The complete MCMC backward-sampling smoother is summarised in Fig.~\ref{alg:MCMC-BSS}.

%\begin{algorithm}
 \begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1$ \TO $N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1$ \TO $1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1$ \TO $M$}
				\STATE Propose a new state hisory, $x_{1:k-1}^{(i)*} \sim \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim q(x_k|x_{k-1}^{*},\tilde{x}_{k+1},y_k)$.
				\STATE With probability $\alpha$ specified by (\ref{eq:MCMC-BSS_ap}),\\ $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
}}
\caption{MCMC backward-sampling smoother algorithm}
\label{alg:MCMC-BSS}
\end{figure}
%\end{algorithm}

@@

The performance of the new smoothers was tested on a simple 2D tracking model. The trajectory of a single target was simulated using linear-Gaussian dynamics and observed via a nonlinear radar-type measurement model (see \cite{Bar-Shalom2002} for details).

%\begin{IEEEeqnarray}{rCl}
%\underbrace{\begin{bmatrix}x_k \\ y_k \\ \dot{x}_k \\ \dot{y}_k\end{bmatrix}}_{\mathbf{x}_k} &=& \underbrace{\begin{bmatrix}I_2 & \Delta t I_2 \\ 0_2 & I_2\end{bmatrix}}_{\mathbf{A}} \underbrace{\begin{bmatrix}x_{k-1} \\ y_{k-1} \\ \dot{x}_{k-1} \\ \dot{y}_{k-1}\end{bmatrix}}_{\mathbf{x}_{k-1}} + \underbrace{\begin{bmatrix}w_{1,k} \\ w_{2,k} \\ w_{3,k} \\ w_{4,k}\end{bmatrix}}_{\mathbf{w}_k} \\
%\underbrace{\begin{bmatrix}b_k \\ r_x\end{bmatrix}}_{\mathbf{y}_k} &=& \begin{bmatrix}\tan^{-1}(y_k / x_k) \\ \sqrt{x_k^2+y_k^2} \end{bmatrix} + \underbrace{\begin{bmatrix}v_{1,k} \\ v_{2,k}\end{bmatrix}}_{\mathbf{v}_k}
%\end{IEEEeqnarray}

\begin{IEEEeqnarray}{rCl}
\underbrace{\begin{bmatrix}x_k \\ y_k \\ \dot{x}_k \\ \dot{y}_k\end{bmatrix}}_{\mathbf{x}_k} &=& \begin{bmatrix}I_2 & \Delta t I_2 \\ 0_2 & I_2\end{bmatrix} \underbrace{\begin{bmatrix}x_{k-1} \\ y_{k-1} \\ \dot{x}_{k-1} \\ \dot{y}_{k-1}\end{bmatrix}}_{\mathbf{x}_{k-1}} + \underbrace{\begin{bmatrix}w_{1,k} \\ w_{2,k} \\ w_{3,k} \\ w_{4,k}\end{bmatrix}}_{\mathbf{w}_k} \\
\underbrace{\begin{bmatrix}b_k \\ r_x\end{bmatrix}}_{\mathbf{y}_k} &=& \begin{bmatrix}\tan^{-1}(y_k / x_k) \\ \sqrt{x_k^2+y_k^2} \end{bmatrix} + \underbrace{\begin{bmatrix}v_{1,k} \\ v_{2,k}\end{bmatrix}}_{\mathbf{v}_k}
\end{IEEEeqnarray}

The time between successive time steps is $\Delta t$. $I_2$ and $0_2$ denote the $2 \times 2$ identity and zero matrices. The random variables $\mathbf{w}_k$ and $\mathbf{v}_k$ are independent and are zero-mean Gaussian distributed with covariance matrices $\mathbf{Q}$ and $\mathbf{R}$ respectively.

@@

The algorithms were all used to generate 100 particles from the joint smoothing distribution, using particle filters with 100 particles. Where appropriate, proposal weights were chosen to be equal to filtering weights, i.e. $v_k^{(j)} = w_k^{(j)}$. The particle filter and MCMC-BSS both use ``optimal importance distributions'' with linearisations of the observation model where appropriate.

10 realisations of 500 time steps were simulated from the model and the algorithms tested on each one. Results are shown here for high and low observation noise cases. For case 1, the observation variances were set to $\sigma_B = \pi/720$ and $\sigma_R = 0.1$. For case 2, $\sigma_B = \pi/36$ and $\sigma_R = 100$. Other model settings are $\Delta t = 1$, $\sigma_P = 1$, and $x_0 = [-100, 50, 10, 0]^T$. Example trajectories and observations from the two cases are shown in figure~\ref{fig:example_trajectories}.

@@

Tables~\ref{tab:case1_performance} and~\ref{tab:case2_performance} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms with the two parameter cases, along with running times (excluding the time for filtering) and the mean number of unique particles at each time instant. The trade-off between speed and error is shown in figures~\ref{fig:case1_rmse_vs_time} and~\ref{fig:case2_rmse_vs_time}.

@@

The MCMC-BRS smoother error approaches that of the DBRS as the number of MCMC steps, $M$, increases. However, for small $M$ the error is only very slightly larger than that of the DBRS but with significantly less running time. For a particular application, it is possible to adjust $M$ to control the trade-off between error and speed.

@@

\section{Conclusion} \label{sec:conclusions}
Two new algorithms have been presented for drawing particles from the joint smoothing distribution using MCMC. These MCMC schemes have a degree of flexibility in the choice of chain length, which controls a trade-off between accuracy and running time.

@@

The second algorithm samples new values of the state from a continuous proposal density. Although slower than the resampling version for a given length of Markov chain, the error performance is reduced relative to the other smoothers.
